---
title: "Lab10"
author: "Matthew"
date: "2/11/2023"
output: html_document
---

First 

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names=1)
head(candy)

```

# Q1. How many different candy types are in this dataset?
```{r}
dim(candy)
```
There are 85 different candy types 

# Q2. How many fruity candy types are in the dataset?
dim(candy)
```{r}
sum(candy$fruity,na.rm=TRUE)
```
There are 38 fruity candy types in the database. 


```{r}
candy["Twix", ]$winpercent
```

# Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Sour Patch Kids", ]$winpercent
```

My favorite candy in the dataset is Sour Patch Kids and it has a winpercent value of 59.864%. 

# Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

Kit Kat has a win percent of 76.7686%. 

# Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

Tootsie Roll Snack Bars has a win percent of 49.6535%. 


```{r}
library("skimr")
skim(candy)
```

# Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

p100 because compared to all the other p value, that value has most of the values close to 1 while the other has most values with values that are 0.0. 

# Q7. What do you think a zero and one represent for the candy$chocolate column?

If it has chocolate or not in the candy.

# Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


# Q9. Is the distribution of winpercent values symmetrical?

No, it seems like the values are more skewed to the left 

# Q10. Is the center of the distribution above or below 50%?

The center of distribution seem to be below 50%. 

# Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc_per <- candy$winpercent[as.logical(candy$chocolate)]
choc_per
```
```{r}
mean(choc_per)
```
```{r}
fruit_per <- candy$winpercent[as.logical(candy$fruity)]
fruit_per
```
```{r}
mean(fruit_per)
```
Chocolate on average is higher ranked than fruity candy. 

# Q12. Is this difference statistically significant?

```{r}
t.test(choc_per,fruit_per)
```
It is statistacally significant since the p value is 2.9e-08, which is much lower than 0.05, which is usually the accepted value of when data is statistcally significant. 


# Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=15)
```

The five least liked cany types in the set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

# Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
tail(candy[order(candy$winpercent),], n=15)
```
The top 5 all time favorite candy types are Snickers, Kit Kat, Twix, Reese's Minatures, and Reese's Peanut Butter cup 


# Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
```

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy),winpercent) +
  geom_col() 
```

# Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() 
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

#  Q17. What is the worst ranked chocolate candy?

Sixlets

#  Q18. What is the best ranked fruity candy?

Starburst


```{r}
library("ggrepel")
```

```{r}
# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

# Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Minature since it has over 80% winpercent with a little over 25% pricepercent. 

# Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


The top 5 most expensive candies are Nik L Lip, Ring pop, Nestl Smarties, Hershey Krackel, and Hersheys Milk Chocolate. The most unpopular candy of these are Nik L Nip 


# Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```




```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

# Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and Chocolate

# Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar or chocolate and winpercent. 



```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2],col=my_cols,pch=16)
```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```



```{r}
library(plotly)
ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

# Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The original variables that are picked up strongly by PC1 in positive direction are fruity, hard, and pluribus. 
