---
title: "Lab11 Session"
author: "Matthew"
date: "2/14/2023"
output: github_document
---

# Section 1, Proportion og G/G in a population 

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39869097-39921098;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel

Here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

Now let's look at a different population. I picked the GBR

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population  than the MKL population 

Now lets dig into this further. 



# Section 4: Population Scale Analysis 


How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```


```{r}
library(ggplot2)
```

Let's make a boxplot 

```{r}
ggplot(expr)+aes(x=geno,y=exp,fill=geno)+
  geom_boxplot(notch=TRUE)
```

# Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

GG shows that it is associated with reduced expression of the gene compared to AA. This shows that the SNP more likely affect the expression of ORMDL3. 



